\textit{MSF: Modulated Sub-graph Finder}

### 2018-05-15 MSF_Paper_FA.tex
last editing

High throughput sequencing techniques have been widely used to yield
differentially expressed genes (DEG)~\cite{DEG}. To this end, changes in
transcript abundance are measured, e.g.~by next generation sequencing
techniques, and interpreted as an indicator of differential expression of
genes. DEGs can be used to get insights into the mechanism underlying
differences between conditions of samples, such as healthy versus
diseased. Differential gene expression analysis (DGEA) informs about the
magnitude of expression changes between the conditions which are often
expressed as fold change, sign of fold change and the confidence level of
observing an authentic change, often expressed as \textit{p}-value. These
DEGs information is further interpreted to extract meaningful biological
insights. For example, genes that could be involved in the response to a
particular stimuli or maybe the cause of a disease. To this end,
pathway-based analysis has become an important tool to further interpret
the results of a DGEA and to acquire understandings of the perturbations in
a biological system. Biological pathways are sets of genes and their
interactions forming a functional unit. DEGs help to identify pathways or
networks that may be altered during a change of condition providing
important information about diseases and its treatment
process~\cite{Khatri2012}. Pathway-based methods use predefined pathways or
networks such as KEGG~\cite{Kegg} and Reactome~\cite{Reactome}, the
expression measurements of the genes obtained from DGEA in combination with
statistical methods and algorithms to identify specifically modulated
pathways and processes~\cite{Campos}.

Well established resources for pathway annotation are KEGG (Kyoto
Encyclopedia of Genes and Genomes)\cite{Kegg} and
Reactome~\cite{Reactome}. KEGG pathways is a branch of KEGG database that
hosts a collection of manually drawn pathway maps representing the
molecular interaction, reaction and relation networks of cellular
functions. Similarly, Reactome is an open-source, manually curated,
peer-reviewed database for signaling and metabolic molecules with their
interactions formed into different biological pathways for nineteen
species~\cite{Reactome}. Both provide predefined pathways which are sets of
genes and their interactions categorized into functional units. Starting
from a gene interaction network, genes are labeled according their role in
a specific biological process. In this sense a particular gene can be
assigned to different pathways. E.g., the human gene STAT1 is associated
with 24 different pathways in the pathway annotation curated by
KEGG~\footnote{\url{http://www.genome.jp/dbget-bin/www_bget?hsa:6772}} and
in 12 different pathways in the Reactome data
base~\footnote{\url{http://www.reactome.org/content/detail/R-HSA-629622}}.
Although carefully produced, the assignment of genes to those predefined
pathway unites can be considered to be subjective to some degree and
suffers from observational bias~\cite{schnoes2013biases}.

Existing pathway-based analysis approaches use different research designs,
which can be categorized into ORA (Over-representation analysis), FCS
(Functional class scoring) and pathway topology based methods. All of which
aim in finding in a subset of genes, e.g., significantly differentially
expressed genes, genes associated with a certain pathway more often than
expected given the total set of examined genes, e.g.~the whole genome
background.  ORA is the first and the most basic method of pathway
analysis. It uses a DEG list with user defined cut-off for the log-fold
change and \textit{p}-value (most commonly using absolute log-fold change
$\geq$ 2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes
associated with annotated pathways are tested for being over-represented in
the set of DEGs. To this end, hyper-geometric distribution, chi-square
tests, binomial probability or the Fisher’s exact test are used. Thereby
the information of the topology of genes in the pathways is
neglected~\cite{Bayer}. Furthermore, ORA assumes that the biological
pathways are independent of each other and ignores the fact that they
cross-talk and overlap~\cite{Khatri2012,Campos}.

Unlike ORA, FCS has no artificial cut-off to define DEG list. FCS works in
three step, first it calculates the gene-level statistics including
correlation of molecular measurements using differential expression of
individual genes, ANOVA, t-test and Z-score. In the second step the
statistics of individual genes in a pathway are transformed to an
individual pathway-level statistic commonly using Kolmogorov-Smirnov
statistic, mean or median. Finally the statistical significance of the
pathway-level statistics is assessed. Although FCS covers some of the
limitations from ORA, it still lacks the topology of genes in a pathway,
cross-talk and overlap of the pathways~\cite{Khatri2012,Campos}. Pathway
topology based methods are similar to FCS except that they consider the
topology of each gene during the gene-level statistics but still don't aim
to link different functional pathways~\cite{Khatri2012}.

@@

As input, information on functional links between genes provided by
e.g.~KEGG or Reactome and information on the differential expression status
of single genes resulting from a DGEA, are required. As a result the
analysis returns sub-graphs and their joint confidence scores, reflecting
how the perturbation is migrated through the network. Furthermore, the
entry points of perturbation in the networks and overlap with conventional
pathway categories are returned. The output is prepared in a directed
adjacency file, convenient for visualization, e.g., with
StringApp~\cite{StringApp}, available as a Cytoscape plug-in~\cite{Cyto}.

@@

In a last post processing step \texttt{MSF} identifies the trigger points
of the modulated sub-graphs. These trigger genes are the sources of the
sub-graphs with only outgoing edges. These genes can be interpreted as the
possible entry points of perturbation from where the stimulus causes
downstream effects. In the same spirit the most downstream genes of the
modulated sub-graph are identified and defined as sinks. Sinks can be
interpreted as the effectors where the integrated information within the
signal transduction network is set to action. Due to circular loops not all
sub-graphs are guaranteed to have sources or sinks.

@@

To demonstrate its usefulness, \texttt{MSF} is applied to an RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test the approach because it has been well recognized for the last several decades, and vast literature is available for the pathogenesis of Ebola. Thereby facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
rightness and usefulness of the approach.

@@

Table~\ref{tab:rawVsHartung} compares the results of \texttt{MSF}, namely
the number of detected sub-modules and their total genes numbers, to a
simple analysis of mapping significantly modulated genes from the DGEA to
the network and joining neighbors to modules. The numbers indicate that
\texttt{MSF} detects less but larger sub-modules, applying its statistical
test.  Furthermore, the dependency of the results from the p-value cutoff
choice is demonstrated for the DGEA, which is avoided for \texttt{MSF}
altogether.

@@

 To assess the robustness and stability of our method, Poisson distributed
 noise was added to the read counts of the three time-points data set, used
 in the previous analysis. Then DGEA was carried out on the disturbed data
 with the same parameters as for the native data using edgeR, followed by
 analysis with \texttt{MSF}. This procedure was carried out 100 times.
 Every time the genes from the modulated sub-graphs identified from noisy
 data were compared to the genes of sub-graphs identified from the native
 data. This was also done for the DEG obtained for each run, using three different cutoffs of
 FDR 0.01, 0.05 and 0.1. The robustness of \texttt{MSF} and the DEG analysis
 for the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
 considered as rather stringent, which is already reflected by the limited
 recall rate in the edgeR based DGEA, between 68~\% (6~hpi) and 93~\%
 (24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
 between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
 pure DGEA can be explained by the fact that disturbed p-values do not
 change the results for DGEA as long as the p-value does not wise above the
 chosen cutoff value. In contrast, \texttt{MSF} is sensitive to p-value
 changes across the whole range of possible values.

@@

Gene enrichment analysis was performed using Reactome Analyze data
tool~\cite{Reactome}. Reactome's over-representation analysis tests whether
certain Reactome pathways are enriched by the list of genes submitted.Genes from \texttt{MSF} identified
sub-graphs for each time-point were analyzed for gene enrichment analysis.  For
comparison the DEG results from edgeR were filtered using three different cut off of
adjusted p-value 0.01, 0.05 and 0.1. This three subsets of DEG list were used for gene
enrichment analysis. The compression of \texttt{MSF}  and the three subset DEG list is shown in Fig~\ref{fig:msfvsreactome}

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{10 cm}
	\includegraphics[width=1.0\linewidth]{MSFvsReactomeUpsetPlot}
	\caption{The Upset plot shows the number of common pathways between different time-point and cut offs. }
	\label{fig:msfvsreactome}
\end{minipage}
}
\end{figure*}

>>>>

High throughput sequencing techniques have been widely used to yield
differentially expressed genes (DEG)~\cite{DEG}. To this end, changes in
transcript abundance are measured, e.g.~by next generation sequencing
techniques, and interpreted as an indicator of differential expression of
genes. DEGs can be used to get insights into the mechanism underlying
differences between conditions of samples, such as healthy versus
diseased. Differential gene expression analysis (DGEA) informs about the
magnitude of expression changes between the conditions which are often
expressed as fold change, sign of fold change and the confidence level of
observing an authentic change, often expressed as \textit{p}-value. These
DEGs information is further interpreted to extract meaningful biological
insights. For example, genes that could be involved in the response to a
particular stimuli or maybe the cause of a disease. To this end,
pathway-based analysis has become an important tool to further interpret
the results of a DGEA and to acquire understandings of the perturbations in
a biological system. Biological pathways are sets of genes and their
interactions forming a functional unit. DEGs help to identify pathways or
networks that may be altered during a change of condition providing
important information about diseases and its treatment
process~\cite{Khatri2012}. Pathway-based methods use predefined pathways or
networks such as KEGG~\cite{Kegg} and Reactome~\cite{Reactome2}, the
expression measurements of the genes obtained from DGEA in combination with
statistical methods and algorithms to identify specifically modulated
pathways and processes~\cite{Campos}.

Well established resources for pathway annotation are KEGG (Kyoto
Encyclopedia of Genes and Genomes)\cite{Kegg} and
Reactome~\cite{Reactome2}. KEGG pathways is a branch of KEGG database that
hosts a collection of manually drawn pathway maps representing the
molecular interaction, reaction and relation networks of cellular
functions. Similarly, Reactome is an open-source, manually curated,
peer-reviewed database for signaling and metabolic molecules with their
interactions formed into different biological pathways for nineteen
species~\cite{Reactome2}. Both provide predefined pathways which are sets of
genes and their interactions categorized into functional units. Starting
from a gene interaction network, genes are labeled according their role in
a specific biological process. In this sense a particular gene can be
assigned to different pathways. E.g., the human gene STAT1 is associated
with 24 different pathways in the pathway annotation curated by
KEGG~\footnote{\url{http://www.genome.jp/dbget-bin/www_bget?hsa:6772}} and
in 12 different pathways in the Reactome data
base~\footnote{\url{http://www.reactome.org/content/detail/R-HSA-629622}}.
Although carefully produced, the assignment of genes to those predefined
pathway unites can be considered to be subjective to some degree and
suffers from observational bias~\cite{schnoes2013biases}.

Existing pathway-based analysis approaches use different research designs,
which can be categorized into ORA (Over-representation analysis), FCS
(Functional class scoring) and pathway topology based methods. All of which
aim in finding in a subset of genes, i.e., significantly differentially
expressed genes, genes associated with a certain pathway more often than
expected given the total set of examined genes, e.g.~the whole genome
background.  ORA is the first and the most basic method of pathway
analysis. It uses a DEG list with user defined cut-off for the log-fold
change and \textit{p}-value (most commonly using absolute log-fold change
$\geq$ 2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes
associated with annotated pathways are tested for being over-represented in
the set of DEGs. To this end, hyper-geometric distribution, chi-square
tests, binomial probability or the Fisher’s exact test are used~\cite{Bayer}. Thereby
the information of the topology of genes in the pathways is
neglected. Furthermore, ORA assumes that the biological
pathways are independent of each other and ignores the fact that they
cross-talk and overlap~\cite{Khatri2012,Campos}.

Unlike ORA, FCS has no artificial cut-off to define DEG list. FCS works in
three step, first it calculates the gene-level statistics including
correlation of molecular measurements using differential expression of
individual genes, ANOVA, t-test and Z-score. In the second step the
statistics of individual genes in a pathway are transformed to an
individual pathway-level statistic commonly using Kolmogorov-Smirnov
statistic, mean or median. Finally the statistical significance of the
pathway-level statistics is assessed. A popular implementation of FCS is \texttt{GSEA}~\cite{subramanian2005gene}.
Although FCS covers some of the
limitations from ORA, it still lacks the topology of genes in a pathway,
cross-talk and overlap of the pathways~\cite{Khatri2012,Campos}. Pathway
topology based methods are similar to FCS except that they consider the
topology of each gene during the gene-level statistics but still don't aim
to link different functional pathways~\cite{Khatri2012}.

@@

As input, information on functional links between genes provided by
e.g.~KEGG or Reactome and information on the differential expression status
of single genes resulting from a DGEA, are required. As a result the
analysis returns sub-graphs and their joint confidence scores, reflecting
how the perturbation is migrating through the network. Furthermore, the
entry points of perturbation in the networks and overlap with conventional
pathway categories are returned. The output is prepared in a directed
adjacency file, convenient for visualization, e.g., with
StringApp~\cite{StringApp}, which is available as a Cytoscape plug-in~\cite{Cyto}.

@@

In a last post processing step \texttt{MSF} identifies the trigger points
of the modulated sub-graphs. These trigger genes are the sources of the
sub-graphs with only outgoing edges. These genes can be interpreted as the
possible entry points of perturbation from where the stimulus causes
downstream effects. In the same spirit the most downstream genes of the
modulated sub-graph are identified and referred to as sinks. Sinks can be
interpreted as the effectors where the integrated information within the
signal transduction network is set to action. Due to circular loops not all
sub-graphs are guaranteed to have sources or sinks.

@@

To demonstrate its usefulness, \texttt{MSF} is applied to an RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test our novel method because it has been well recognized for the last several decades, and vast literature is available for the pathogenesis of Ebola. Thereby facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
rightness and usefulness of the approach.

@@

Table~\ref{tab:rawVsHartung} compares the results of \texttt{MSF}, namely
the number of detected sub-modules and their total gene numbers, to a
simple analysis of mapping significantly modulated genes from the DGEA to
the network and joining neighbors to modules. The numbers indicate that
\texttt{MSF} detects less but larger sub-modules, applying its statistical
test.  Furthermore, the dependency of the results from the p-value cutoff
choice is demonstrated for the DGEA, which is avoided for \texttt{MSF}
altogether.

@@

 To assess the robustness and stability of our method, Poisson distributed
 noise was added to the read counts of the three time-points data set, used
 in the previous analysis. Then DGEA was carried out on the disturbed data
 with the same parameters as for the native data using edgeR, followed by
 analysis with \texttt{MSF}. This procedure was carried out 100 times.
 Every time the genes from the modulated sub-graphs identified from noisy
 data were compared to the genes of sub-graphs identified from the native
 data. This was also done for the DEG obtained for each run, using three different cutoffs of
 FDR 0.01, 0.05 and 0.1. The robustness of \texttt{MSF} and the DEG analysis
 for the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
 considered as rather stringent, which is already reflected by the limited
 recall rate in the edgeR based DGEA, between 68~\% (6~hpi) and 93~\%
 (24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
 between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
 pure DGEA can be explained by the fact that disturbed p-values do not
 change the results for DGEA as long as the p-value does not rise above the
 chosen cutoff value. In contrast, \texttt{MSF} is sensitive to p-value
 changes across the whole range of possible values.

@@

Gene enrichment analysis was performed using Reactome Analyze data
tool~\cite{Reactome2}. Reactome's over-representation analysis tests whether
certain Reactome pathways are enriched by the list of genes submitted. Genes from \texttt{MSF} identified
sub-graphs for each time-point were analyzed for gene enrichment analysis.  For
comparison the DEG results from edgeR were filtered using three different cut off of
adjusted p-value 0.01, 0.05 and 0.1. This three subsets of DEG list were used for gene
enrichment analysis. The compression of \texttt{MSF}  and the three subset DEG list is shown in Fig~\ref{fig:msfvsreactome}

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{15 cm}
	\includegraphics[width=1.0\linewidth]{MSFvsReactomeUpsetPlot}
	\caption{The Upset plot shows the number of common pathways between different time-point and cut offs. }
	\label{fig:msfvsreactome}
\end{minipage}
}
\end{figure*}
